---
title: "Log of Crab body metrics - predicting the subspecies of crabs"
author: "IJsbrand Pool, 403589"
output:
  pdf_document:
    toc: true 
    fig_caption: yes
    fig_crop: no
    highlight: haddock
    keep_tex: yes
    number_sections: yes
header-includes: 
- \usepackage{graphicx}
- \usepackage{float}
linkcolor: blue
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      fig.align='center',
                      external=TRUE,
                      warning=FALSE,
                      fig.pos='H')
```

```{r libraries, message=F}
library(ggplot2)
library(factoextra)
library(FactoMineR)
library(RWeka)
library(reshape)
library(kableExtra)
library(forcats)
library(dplyr)
library(ggcorrplot)
library(gridExtra)
library(grid)
library(ggpubr)
```

\newpage

# Logbook EDA

## Dataset information
The rock crab *Leptograpsus variegatus*, is recorded as occurring on a number of southern Pacific islands, the western coast of South America, and the coasts of Australia south of the Tropic of Capricorn. Mahon, using ecological studies which extended those of Shield, and a genetical analysis based on an electrophoretic study, established the specific distinctness of rock crabs of the blue and orange forms of the genus *Leptograpsus* which occur on the coasts of Australia. These colour forms were previously regarded as morphs of *L. variegatus.* 

In an attempt to resolve this problem of identification, a morphological study of the Western Australian species was undertaken. This paper reports an exploratory data analysis of the data and a machine learning algorithm to predict the species of the crab based on this data.

The dataset used is the crab body metrics dataset by Campbell, N.A. and Mahon, R.J. (1974) A multivariate study of variation in two species of rock crab of genus *Leptograpsus*[1]. This data set contains multiple morphological metrics of the bodies of *L. variegatus.* crabs, and the gender and color of the crab. The measured metrics are the frontal lobe size, rear width, carapace length, carapace width and body depth. All these values are in millimeters. There are 100 orange and 100 blue crabs. 50 crabs of each gender per color of crab.


### Research question
The goal of this project was to find out if the subspecies of a *Leptograpsus variegatus* was predictable.To find out if this is possible, a research question had to be formulated. The research question for this project is "Can the species of a *L.variegatus* be determined based on some morphological measurements of its carapace". To answer this question, the data was first explored and cleaned. Then, multiple machine learning algorithms were tested to find what algorithm could be used best.

\newpage

## Data analysis
```{r Reading data, echo=FALSE}
# Load in the data
myData <- read.csv("datafiles/data.csv")

# Making the two separate subgroups
bluecrabs <- myData[myData$sp == "B",]
orangecrabs <- myData[myData$sp == "O",]

# Creating the code book
column <- colnames(myData)
description <- c("Species", "Sex", "Index", "Frontal lobe size (mm)", "Rear width (mm)", "Carapace length (mm)", "Carapace width (mm)", "Body depth (mm)")
codebook <- data.frame(column, description)

kable(codebook, caption = "Codebook of the crab data set") %>% 
kable_styling(font_size = 10, latex_options = "hold_position")
```
After the data was loaded in, a codebook with the attribute names and their descriptions was generated, shown in table 2. To get a better view of the measurements in the data set, a five number summary was created for each attribute. These values are shown in table 3. The table shows that the mean and the median are close in value for each column, meaning that they all are normal distributions.


```{r summary}
sumdat <- summary(myData[4:8])
sumdat <- sub(".*:", "", sumdat)
rownames(sumdat) <-  c("Minimum", "Q1", "Median", "Mean", "Q3", "Maximum")
colnames(sumdat) <- c("Frontal Lobe", "Rear Width", "Carapace Length", "Carapace Width", "Body Depth")
kable(sumdat, caption = "Five number summary of the morphological measurements of the crab bodies.") %>% 
kable_styling(latex_options = "hold_position")
```

The data in its current form, is not ready for the use of a machine learning algorithm, this is due to the Index column, some algorithms will over fit their model by making use of this column, so this column needs to be deleted to be able to verify the quality of the data gathered. The species columns was also moved to the last column so it would be used as the class attribute.

```{r Data cleaning}
# Cleaning the data by the removal of column Index.
clean_data <- myData[,c(2,4:ncol(myData), 1)]
#write.csv(clean_data, "datafiles/cleanedData.csv", row.names = F, col.names = F)
```

\newpage

## Visualisation
To gain a better understanding of the correlation between the attributes in the data set, some visualizations were made.

```{r figure1, fig.cap= "Spread of Front lobe size against Carapace width based on color", fig.height = 3.5, fig.width = 5, warning=F}
#plot the Front lobe size against Carapace width
ggplot(data = clean_data,  
aes(x = FL, y = CW)) + 
  geom_jitter(alpha = 0.5, aes(colour = factor(sp))) +  
  geom_smooth(method = "lm", se = FALSE, formula="y~x", aes(colour =   factor(sp))) +
  scale_color_manual(values=c("Blue","Orange")) +
  ggtitle("Carapace width vs Frontal lobe size") + 
  xlab("Frontal Lobe Size (mm)") + ylab("Carapace Width (mm)") +
  scale_colour_discrete(name="Crab species")
```
The image above shows the spread of front lobe size against Carapace width based on color, this visualization has been made to be able to see if there was any correlation between these two attributes. The image shows that the blue crabs have wider carapaces on average and shorter frontal lobes, this could be a genetic trait for the color of this crab and therefore this could be a good indicator to determine the subspecies of the crab by the use of a machine learning algorithm. 

Since there were only 200 crabs in this study this could of course be a coincidence, but for the continuation of this report the assumption of data correctness due to differences in species has been accepted.

```{r figure2, fig.cap= "Spread of Rear width against Carapace length based on color", fig.height = 3.5, fig.width = 5}
#plot the Rear width against Carapace length
ggplot() +
  geom_point(data = myData[myData$sp == "B",], mapping = aes(x = RW, y = CL, color = 'Blue')) +
  geom_point(data = myData[myData$sp == "O",], mapping = aes(x = RW, y = CL, color = 'Orange')) +
  scale_color_manual(values=c("#4444EE", "#E69F00")) +
  labs(x = "Rear width (mm)", y = "Carapace length (mm)", title='Rear width against Carapace length')
```
Similar to the previous plot, this data also seems to show some correlation with each other, this makes the question "is all the data correlated" rise, this will have to be investigated if the data continues to seem correlated. In the plot above the differences between Rear width against Carapace length are visible for both the different crabs, this being the blue group and the orange group.

\newpage
```{r figure3, fig.cap="Spread of Body depth against Carapace width based on color", fig.height = 3.5, fig.width = 5}
#plot the Body depth against Carapace width
ggplot() +
  geom_point(data = myData[myData$sp == "B",], mapping = aes(x = BD, y = CW, color = 'Blue')) +
  geom_point(data = myData[myData$sp == "O",], mapping = aes(x = BD, y = CW, color = 'Orange')) +
  scale_color_manual(values=c("#4444EE", "#E69F00")) +
  labs(x = "Body depth (mm)", y = "Carapace width (mm)", title='Body depth against Carapace width')
```
This plot again shows that the blue crabs on average seem to have a wider carapaces, but the orange crabs tend to have deeper bodies. The attributes seem to very correlated with each other and this again supports the theory that there may be multiple morphological attributes that will help determine the color of the crab. However the attributes above seem to be less clear in this then the attributes within figure one.

As shown inside the figures one and three, the metrics of the orange and the blue crabs appear to be in two different groups. This could indicate that the two species could be identifiable by different elements, including their carapace width. However, figure two indicates that the two groups of crabs appear to have some sort of difference even within their own respective group, this could indicate that gender plays a role in the sizes of the morphological measurements.

\newpage
```{r figure4, fig.cap="Checking if the gender makes a difference"}
library(ggpubr)

densityplot1 <- ggplot(clean_data, aes(FL), alpha = 0.5) + geom_density(aes(col = sex, fill=sex), alpha = 0.5) +
ggtitle("FL") + ylab("Density")


densityplot2 <- ggplot(clean_data, aes(RW), alpha = 0.5) + geom_density(aes(col = sex, fill=sex), alpha = 0.5) +
ggtitle("RW") + ylab("Density")

densityplot3 <- ggplot(clean_data, aes(CL), alpha = 0.5) + geom_density(aes(col = sex, fill= sex), alpha = 0.5) +
ggtitle("CL") + ylab("Density")

densityplot4 <- ggplot(clean_data, aes(CW), alpha = 0.5) + geom_density(aes(col = sex, fill=sex), alpha = 0.5) +
ggtitle("CW") + ylab("Density")

densityplot5 <- ggplot(clean_data, aes(BD), alpha = 0.5) + geom_density(aes(col = sex,fill= sex), alpha = 0.5) +
ggtitle("BD") + ylab("Density")

densityplots <- ggarrange(densityplot1, densityplot2, densityplot3, densityplot4, densityplot5,
                   ncol = 2, nrow = 3)

title <- expression(atop(bold("Expression data"), 
                         scriptstyle(paste("Density plots of all the morhpological measurements compared to sex"))))
annotate_figure(densityplots,
                top=text_grob(title))
```
As seen in the density distribution of figure four, there indeed seems to be a difference within the groups if they are divided by the male and females of the group. 

In figure 4.1 the frontal lobe size in millimeters is plotted against the gender of the crabs, here it is clearly visible that there doesn't seem to be much difference in the frontal lobe size between the genders, the only difference that can be seen here appears to be in the the smaller measurements of the frontal lobe size, slightly indicating that the females might have a smaller frontal lobe size in general then males, this however cannot be confirmed from this density plot.

In figure 4.2 the Rear width in millimeters is plotted against the gender of the crabs, here it can be seen that the female crabs appear to have a bigger rear width then the male crabs, it could be that one colour species has a larger rear width, but the most plausible reason for this is that genetically speaking the female crab has a larger rear width.

In figure 4.3 the Carapace length in millimeters is shown here against the gender of the crabs, here similar to figure 4.1 there do not seem to be many differences apart from the fact that more females appear to have a carapace length of around 30 รก 35 millimeters in length, this however does not seem significant, and that a small portion of the male crabs seem to have a larger carapace length then the females, but since this doesn't seem like a mayority this could just be an anomaly.

In figure 4.4 the Carapace width is shown, measured in millimeters, plotted against the gender of the crabs similar to the figure 4.3 there do not seem to be a lot of differences, apart from the males where a small portion seem to have a larger carapace width then the females, this is however again a small portion so it may be insignificant.

In figure 4.5 the Body depth in millimeters is shown, this is plotted against the gender of the crabs, here the distributions appear similar, making is seem like gender does not affect body depth what so ever.

One thing that is noticeable here is the distributions of the female crabs, these appear to be more of a normal distribution then the male crabs, which can be due to coincidence since it is a small data set, but it can also be that the body of a female crab is more normally distributed then the body of its male counterpart. 

```{r}
corr <- round(cor(clean_data[2:6]), 2)

# Plot
ggcorrplot(corr, hc.order = TRUE, 
           type = "lower", 
           lab = TRUE, 
           lab_size = 3, 
           method="circle", 
           colors = c("tomato2", "white", "springgreen3"), 
           title="Correlogram of the crabs", 
           ggtheme=theme_bw)
```
In the plot above the correlation between all the numerical elements of the data set are shown, this has been done to be able to answer the previously arisen question whether or not the data is correlated.

As seen in the plot above, the data inside the data set does indeed seem to correlate which supports the earlier findings, this could mean that there is more then one good attribute to be able to confidently support the investigation of the research question


```{r figure6, fig.cap="Checking if the gender makes a difference"}
orangedensityplot1 <- ggplot(orangecrabs, aes(FL), alpha = 0.5) + geom_density(aes(col = sex, fill=sex), alpha = 0.5) +
ggtitle("FL") + ylab("Density")

orangedensityplot2 <- ggplot(orangecrabs, aes(RW), alpha = 0.5) + geom_density(aes(col = sex, fill=sex), alpha = 0.5) +
ggtitle("RW") + ylab("Density")

orangedensityplot3 <- ggplot(orangecrabs, aes(CL), alpha = 0.5) + geom_density(aes(col = sex, fill= sex), alpha = 0.5) +
ggtitle("CL") + ylab("Density")

orangedensityplot4 <- ggplot(orangecrabs, aes(CW), alpha = 0.5) + geom_density(aes(col = sex, fill=sex), alpha = 0.5) +
ggtitle("CW") + ylab("Density")

orangedensityplot5 <- ggplot(orangecrabs, aes(BD), alpha = 0.5) + geom_density(aes(col = sex,fill= sex), alpha = 0.5) +
ggtitle("BD") + ylab("Density")

orangedensityplots <- ggarrange(orangedensityplot1, orangedensityplot2, orangedensityplot3, orangedensityplot4, orangedensityplot5,
                   ncol = 2, nrow = 3)

title <- expression(atop(bold("Expression data"), 
                         scriptstyle(paste("Density plots of all the morhpological measurements compared to sex"))))
annotate_figure(orangedensityplots,
                top=text_grob(title))
```
As seen in figure six, the orange crabs their morphological measurements have been taken to be able to see if the sex matters when it comes to the colours of the crabs, 
```{r figure 7ALSE}

bluedensityplot1 <- ggplot(bluecrabs, aes(FL), alpha = 0.5) + geom_density(aes(col = sex, fill=sex), alpha = 0.5) +
ggtitle("FL") + ylab("Density")

bluedensityplot2 <- ggplot(bluecrabs, aes(RW), alpha = 0.5) + geom_density(aes(col = sex, fill=sex), alpha = 0.5) +
ggtitle("RW") + ylab("Density")

bluedensityplot3 <- ggplot(bluecrabs, aes(CL), alpha = 0.5) + geom_density(aes(col = sex, fill= sex), alpha = 0.5) +
ggtitle("CL") + ylab("Density")

bluedensityplot4 <- ggplot(bluecrabs, aes(CW), alpha = 0.5) + geom_density(aes(col = sex, fill=sex), alpha = 0.5) +
ggtitle("CW") + ylab("Density")

bluedensityplot5 <- ggplot(bluecrabs, aes(BD), alpha = 0.5) + geom_density(aes(col = sex,fill= sex), alpha = 0.5) +
ggtitle("BD") + ylab("Density")

bluedensityplots <- ggarrange(bluedensityplot1, bluedensityplot2, bluedensityplot3, bluedensityplot4, bluedensityplot5,
                   ncol = 2, nrow = 3)

title <- expression(atop(bold("Expression data"), 
                         scriptstyle(paste("Density plots of all the morhpological measurements compared to sex"))))
annotate_figure(bluedensityplots,
                top=text_grob(title))
```

In figure six and seven the morphological variables of both color crabs have been put into a density plot to see if there is any correlation there, here it is however clearly visible these two separate colors differ in their measurements. As can be seen in figure six: There are no real differences with the morphological measurements for the orange crabs, apart from the rear width, and here the females are noted to have a larger rear width, apart from that some males appear to be on the lager side of the spectrum in the density plot, but this does not seem significant enough that it is to be mentioned. The orange crabs also seem to have somewhat of a normal distribution or bell curve when it comes to their measurements, this could indicate that they are either smaller or their morphological components are more formed to each other.

In figure seven the blue crabs are depicted, here differing from figure six, the male crabs of the blue species seem to be bigger in almost all aspects of the morphological components apart from rear width, this would indicate that for this species the male variant is bigger then the female variant, how ever, this is a data set with only 100 crabs per species and for a correct distribution a much larger sample size should be taken. For the carapace length, it seems that for this species alone it would be very good to be determined if the crab was of significant size, which species it would belong to. To see if this would work with a machine learning algorithm, further testing will need to be done.



```{r pca, fig.height = 3.5, fig.width = 5}
res.pca <- PCA(myData[4:ncol(myData)], ncp = 5, graph = FALSE)
fviz_pca_var(res.pca,
                      col.var = "contrib", # Colour by contributions to the PC
                      gradient.cols = c("orange", "purple"),
                      repel = TRUE     # Avoid text overlapping
                      ) 
```

This PCA plot shows that the variables are highly correlated. The least correlated variable is the Rear width. This vector has the largest angle with the Carapace length, the same two variables were used to discover the difference in carapace length distribution between the genders.

## Machine learning
### cleaning
Before machine learning can be used, the data must first be cleaned. In its current form, the data is not ready for machine learning. Because of the id column, some of the machine learning algorithms overfit their model by using this column. So this column should be removed first. The species columns was also moved to the last column so it would be used as the class attribute.
```{r datacleaning}
clean_data <- myData[,c(2,4:ncol(myData), 1)]
write.csv(clean_data, "datafiles/cleanedData.csv", row.names = F, col.names = F)
```

### WEKA
To find out what machine learning algorithm is the best for predicting the species of crab, multiple algorithms were tested. These algorithms are ZeroR, OneR, Simple logistic, Naive bayes, Random forest, J48, SMO and K-nearest neighbor. These algorithms were tested using 10 fold cross-validation. The highest quality metric for this dataset is the accuracy, since it does not matter whether a blue crab is predicted to be orange, or an orange crab to be blue. The software used to calculate the accuracy is weka. After the classification, the accuracy of these algorithms was saved in a csv file, and are shown in this barplot below.
```{r ml, fig.cap="The accuracy of the machine learning algorithms ordered from low to high."}
algorithms <- read.csv("datafiles/ml.csv")
algorithms <- data.frame(algorithms)

ggplot(algorithms, aes(x = reorder(Algorithm, desc(Accuracy)), y = Accuracy, fill = Algorithm)) +
  geom_bar(stat="identity", alpha=.6, width=.4) +
  coord_flip() +
  xlab("") +
  theme_bw()
```
This plot shows a few interesting things, most notably the 100% accuracy of Simple logistic. The output in weka of the classification using Simple logistic with 10 fold cross-validation shows a model for each species. The model for the blue crabs shows 1.03 + [FL] * -0.6 + [CW] * 0.31 + [BD] * -0.22. The model for the orange crabs shows -1.03 + [FL] * 0.6 + [CW] * -0.31 + [BD] * 0.22. The values in the model of the orange crabs are the values of the model of the blue crabs times -1. The metrics used in the models are Front lobe size, Carapace width and Body depth. The barplot also shows an exact 50% accuracy for the ZeroR algorithm. This is expected since the data has the same amount of blue crabs as orange crabs.


\newpage


```{r metalearners}
metaalgorithms <- read.csv("datafiles/meta ml.csv")
metaalgorithms <- data.frame(metaalgorithms)

ggplot(metaalgorithms, aes(x = reorder(Algorithm, desc(Accuracy)), y = Accuracy, fill = Algorithm)) +
  geom_bar(stat="identity", alpha=.6, width=.4) +
  coord_flip() +
  xlab("") +
  theme_bw()
```


\newpage

## Discussion and future research
The goal was to get the dataset ready for machine learning. The data was analyzed and cleaned to make so it is ready to be used in machine learning algorithms. The data does not contain many outliers. The data points seem easy to classify since most plots show clear groups of blue and orange crabs. As shown in figure 4, the gender of the crab could also be a good attribute to help predict the species of crab. The data also had to be cleaned. This was done by removing the index column, since this column can not be used to help determine the species of crab. It might also be a problematic attribute for some machine learning algorithms. Then, the species column was moved to the last column, so the machine learning algorithms will use this column as the class index.

Future research can be used to show more correlation between other measurements. It can be researched whether or not the gender of the crab could be predicted using these morphological measurements. Machine learning use could also be improved by expanding the dataset, or getting different amounts of blue or orange crabs, with different amounts of male and female crabs. 

\newpage
